---
title: 'Many North American plant and animal populations are sensitive to climate variability: Figures'
author: "Benedicte Bachelot, Aldo Compagnoni, Pedro F.P.  Brandão-Dias, Shannon K. Carter, Marion L. Donald, Joshua C. Fowler, Daniel Gorczynski, Carsten G.B. Grupstra, Zoey R. Neale, Linyi Zhang, Jennifer A. Rudgers, Kai Zhu and Thomas E. Miller"
date: "July 11, 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(popler)
library(dplyr)
library(plyr)
library(tidyverse)
library(maps)
library(ggplot2)
library(scatterpie)
library(mgcv)
library(usmap)
library(lme4)
library(gridExtra)
library(SPEI)
library(rsq)
library(rWBclimate)
```

## Model fit

We analyzed 45 studies. First, let's see how the models fit the data. We do this quickly by looking at the Bayesian p value. We can also investigate other measures of goodness of fit, but the B p is a good first pass. 

```{r load results, echo=FALSE}
setwd("/path/RESULTS")
studies<-list.files()
RESULTS<-list()
bp<-c()
study_length<-c()
type<-c()
ID<-c()
for(i in 1:length(studies))
{
  load(studies[i])
  command <- paste0('pplr_browse( proj_metadata_key==',output$k,', full_tbl = T)')
  ID<-c(ID,output$k)
  metadat<-parse(n=1, text=command) %>% eval
  type <- c(type,metadat$datatype)
  RESULTS[[i]]<-output
  bp<-c(bp,output$bayes.p)
  study_length<-c(study_length,length(unique(output$r_clim$year)))
}

results<-cbind(studies,bp,study_length,type)
results<-data.frame(results)
print(results)

print(summary(lm(as.numeric(as.character(results$bp))~as.numeric(as.character(results$study_length)))))

hist(as.numeric(results$bp), main="Bayesian p value")

organism<-c("insect","plant","plant","plant","plant","plant","animal","plant","plant","animal","animal","insect","bacteria","insect","mollusc","plant","plant","animal","arthorpods","plant","insect","animal","animal","animal","insect","animal","animal","plant","animal","chormiste","animal","animal","animal","animal","animal","plant","birds","birds","birds","insect","plant","plant","plant","animal","animal","animal")

hist(as.numeric(as.character(results$study_length)),xlab="Study duration",main="",xlim=c(0,65),breaks=20)

```

Overall, the models did a decent job: Mean Bayesian p value of `mean(as.numeric(results$bp))` with a range of `range(as.numeric(results$bp))`.

## Jensen's inequality sign

Let's look at the Jensen's inequality. As a reminder, a negative V means that climate variability has a negative effects of species growth rate. We can look at the V statistics in two ways: across the entire posterior distributions:

```{r V statistics, echo=FALSE}
par(mfrow=c(2,2),mar=c(3,3,3,0))
ss<-as.numeric(rep(results$study_length[1],ncol(RESULTS[[1]]$gam_type)))
V<-RESULTS[[1]]$V
Type<-RESULTS[[1]]$gam_type
K<-rep(as.numeric(RESULTS[[1]][2]),ncol(RESULTS[[1]]$gam_type))
  hist(RESULTS[[1]]$V,main=studies[1],xlab="",ylab="")
  abline(v=0,col="red",lwd=2)
for( i in 2:length(studies))
{ 
  V<-cbind(V,RESULTS[[i]]$V)
  Type<-cbind(Type,RESULTS[[i]]$gam_type)
  K<-c(K,rep(as.numeric(RESULTS[[i]][2]),ncol(RESULTS[[i]]$gam_type)))
  ss<-c(ss,as.numeric(rep(results$study_length[i],ncol(RESULTS[[i]]$gam_type))))
  hist(RESULTS[[i]]$V,main=studies[i],xlab="",ylab="")
  abline(v=0,col="red",lwd=2)
}

  # study length effect of detecting anything across the full distributions of gam
  test<-c(1-length(Type[Type[,1]==0,1])/100)
  for(i in 2:912)
  {
          test<-c(test,(1-length(Type[Type[,i]==0,i])/100))
          }
  str(test)
   v_v_glmer<-glmer(test~scale(ss)+(1|K),family="binomial",weights=rep(100,912))
   summary(v_v_glmer)
   plot(scale(ss),test,pch=20)
   yy<-predict(v_v_glmer,type="response",re.form=NA)
   points(scale(ss),yy,pch=20,col="red",type="l",lwd=2)
   
   #assign species a shape -> 0 or 1 for non-sensitive and sensitive.
   sp_sensitive<-rep(0,ncol(Type))
   newtype<-Type
   for(i in 1:ncol(Type))
   {
   newtype[newtype[,i]!=0,i]<-1
   }
   
     for(i in 1:ncol(newtype))
 {
       if((c(0,1)[summary(factor(newtype[,i],levels=c(0,1)))==max(summary(factor(newtype[,i],levels=c(0,1))))])[1]!=0)
 {sp_sensitive[i]<-1}
     }
      sp_glmer<-glmer(sp_sensitive~ss+(1|K),family="binomial")
   summary(sp_glmer)
   plot(ss,sp_sensitive,pch=20,ylab="Climate sensitive",xlab="Study duration")
   yy<-predict(sp_glmer,type="response",re.form=NA)
   points(ss[order(ss)],yy[order(ss)],pch=20,col="red",type="l",lwd=2)
 
  #test for study length effects on V
  ss<-c()
  for(i in 1:length(K))
  {
    ss<-c(ss,unique(study_length[c(1:46)[ID==K[i]]]))
  }
  v_m<-apply(V,2,median)
  v_v<-apply(V,2,sd)
 v_v_glmer<-glmer(scale(v_v,center=FALSE)~scale(ss)+(1|K),family=Gamma(link = "inverse"))
  v_m_lmer<-lmer(v_m~ss+(1|K))
  summary(v_v_glmer)
  summary(v_m_lmer)
    par(mfrow=c(1,2))
    plot(ss,v_m,xlab="Study length (years)",ylab="Jensen's inequality (median)",pch=20)
    plot(ss,v_v,xlab="Study length (years)",ylab="Jensen's inequality (sd)",pch=20)
  
    # test for study length on certainty of the shape
    certain<-c()
    for (i in 1:length(K))
    {
      certain<-c(certain,max(summary(factor(Type[,i],levels=c(0,1,2,3)))))
    }
    certain<-certain/100
    certain_glmer<-glmer(certain~ss+(1|K),family="binomial",weights=rep(100,912))
       plot(ss,certain,pch=20,ylab="Certainty in the shape",xlab="Study duration")
   yy<-predict(certain_glmer,type="response",re.form=NA)
   points(ss[order(ss)],yy[order(ss)],pch=20,col="red",type="l",lwd=2)
```

Instead of looking at each study individually, we can look at the results across studies.

```{r V statistics all, echo=FALSE}
summary(as.numeric(V))
summary(as.numeric(V[Type!=0]))

#let's look at how many unique species
sp<-c()
 for(i in 1:length(RESULTS))
 {
     sp<-c(sp,RESULTS[[i]]$spp_names)
 } 

sp_per_study<-c()
 for(i in 1:length(RESULTS))
 {
     sp_per_study<-c(sp_per_study,length(unique(RESULTS[[i]]$spp_names)))
 } 
sp_per_study_m<-mean(sp_per_study)
sp_per_study_r<-range(sp_per_study)
print(paste("there is on average",sp_per_study_m," per study with a range of (",sp_per_study_r[1],":",sp_per_study_r[2],")"))

sp_unique<-c()
 for(i in 1:100)
     {
         sp_unique<-c(sp_unique,length(unique(sp[Type[i,]!=0])))
 }


par(mfrow=c(1,2))
hist(V,main="V",xlab="",ylab="Number",breaks=20)
  abline(v=0,col="red",lwd=2)
hist(V[Type!=0],main="Significant V",xlab="",ylab="Number",breaks=20)
  abline(v=0,col="red",lwd=2)
  
 zeros<-c(summary(as.factor(Type[1,]))[1]/ncol(V))
 for(i in 2:100)
 {
 zeros<-c(zeros,summary(as.factor(Type[i,]))[1]/ncol(V))
 }
 

hist(zeros,main="Not significant gams")

ones<-c(summary(as.factor(Type[1,]))[2]/ncol(V))
 for(i in 2:100)
 {
 ones<-c(ones,summary(as.factor(Type[i,]))[2]/ncol(V))
 }

hist(ones, main="Linear gams")

twos<-c(summary(as.factor(Type[1,]))[3]/ncol(V))
 for(i in 2:100)
 {
 twos<-c(twos,summary(as.factor(Type[i,]))[3]/ncol(V))
 }

hist(twos, main="Concav or convex gams")

two_negative<-length(V[1,Type[1,]==2][V[1,Type[1,]==2]<0])/ncol(V)
for(i in 2:100)
{
 two_negative<-c(two_negative,length(V[i,Type[i,]==2][V[i,Type[i,]==2]<0])/ncol(V)) 
}

two_positive<-length(V[1,Type[1,]==2][V[1,Type[1,]==2]>0])/ncol(V)
for(i in 2:100)
{
 two_positive<-c(two_positive,length(V[i,Type[i,]==2][V[i,Type[i,]==2]>0])/ncol(V)) 
}

threes<-c(summary(as.factor(Type[1,]))[4]/ncol(V))
 for(i in 2:100)
 {
 threes<-c(threes,summary(as.factor(Type[i,]))[4]/ncol(V))
 }
hist(threes, main="concave and convex gams")

print(paste("Fraction of species with change no SPEI relationships:",round(median(zeros),2),"(",round(quantile(zeros,0.025),2),":",round(quantile(zeros,0.975),2),")"),sep="")

print(paste("Fraction of species with linear gam:",round(median(ones),2),"(",round(quantile(ones,0.025),2),":",round(quantile(ones,0.975),2),")"),sep="")

print(paste("Fraction of species with concave/convex gam:",round(median(twos),2),"(",round(quantile(twos,0.025),2),":",round(quantile(twos,0.975),2),")"),sep="")

print(paste("Fraction of species with concave:",round(median(two_negative),2),"(",round(quantile(two_negative,0.025),2),":",round(quantile(two_negative,0.975),2),")"),sep="")

print(paste("Fraction of species with convex:",round(median(two_positive),2),"(",round(quantile(two_positive,0.025),2),":",round(quantile(two_positive,0.975),2),")"),sep="")

print(paste("Fraction of species with change in concavity:",round(median(threes),2),"(",round(quantile(threes,0.025),2),":",round(quantile(threes,0.975),2),")"),sep="")

```

Fraction of species with no significant gams: 'round(median(zeros),2)' ('round(quantile(zeros,0.025),2)':'round(quantile(zeros,0.975),2)')

Fraction of species with linear gams: 'round(median(ones),2)' ('round(quantile(ones,0.025),2)':'round(quantile(ones,0.975),2)')

Fraction of species with concave or convex gams: 'round(median(twos),2)' ('round(quantile(twos,0.025),2)':'round(quantile(twos,0.975),2)')

Fraction of species with concave gams: 'round(median(two_negative),2)' ('round(quantile(two_negative,0.025),2)':'round(quantile(two_negative,0.975),2)')

Fraction of species with convex gams: 'round(median(two_positive),2)' ('round(quantile(two_positive,0.025),2)':'round(quantile(two_positive,0.975),2)')

Fraction of species with concave and convex gams: 'round(median(threes),2)' ('round(quantile(threes,0.025),2)':'round(quantile(threes,0.975),2)')

## Results through space

Let's summarize the GAM shape. 
```{r Space summarizing GAM shape, echo=FALSE}
lon<-c()
lat<-c()
ID<-c()
LTER<-c()
gam_type<-c() #GAM proportion
for(i in 1:nrow(results))
{
 ID<-c(ID,RESULTS[[i]]$k)
 lon<-c(lon,RESULTS[[i]]$site_lon) 
 lat<-c(lat,RESULTS[[i]]$site_lat)
 pos_neg<-length(RESULTS[[i]]$V[RESULTS[[i]]$gam_type==2][RESULTS[[i]]$V[RESULTS[[i]]$gam_type==2]<0])/length(RESULTS[[i]]$V[RESULTS[[i]]$gam_type==2])
 pos<-(1-pos_neg)*as.numeric(summary(factor(RESULTS[[i]]$gam_type,levels=c("0","1","2","3")))/sum(summary(as.factor(RESULTS[[i]]$gam_type))))[3]
 neg<-pos_neg*as.numeric(summary(factor(RESULTS[[i]]$gam_type,levels=c("0","1","2","3")))/sum(summary(as.factor(RESULTS[[i]]$gam_type))))[3]
 gam_type<-rbind(gam_type,c(as.numeric(summary(factor(RESULTS[[i]]$gam_type,levels=c("0","1","2","3")))/sum(summary(as.factor(RESULTS[[i]]$gam_type))))[1:2],neg,pos,as.numeric(summary(factor(RESULTS[[i]]$gam_type,levels=c("0","1","2","3")))/sum(summary(as.factor(RESULTS[[i]]$gam_type))))[4]))
 LTER<-c(LTER,RESULTS[[i]]$study_site)
}
locations<-cbind(lon,lat,gam_type,ID,LTER)
locations<-data.frame(locations)
colnames(locations)<-c("lon","lat","horizontal","linear","concave","convex","curved2","ID","LTER")

locations<-data.frame(locations)
for(i in 1:7)
{
  locations[,i]<-as.numeric(as.character(locations[,i]))
  locations[is.na(locations[,i])==TRUE,i]<-0
}


location_transform<-usmap_transform(locations[,1:2])

lon2<-c()
lat2<-c()
 for(i in 1:nrow(locations))
     {
         lon2<-c(lon2,location_transform[location_transform[,1]==locations[i,1],3])
         lat2<-c(lat2,location_transform[location_transform[,2]==locations[i,2],4])
 }
locations<-cbind(locations,lon2,lat2)
####
```

Let's put these results on a map. How do the results look across the US? 
```{r Summarized by locations, echo=FALSE}
locations_summarized<-matrix(data=0,nrow=length(unique(locations$LTER)),ncol=8)
locations_summarized[,8]<-unique(locations$LTER)
for(i in 1:nrow(locations_summarized))
{
  locations_summarized[i,1:7]<-apply(locations[locations[,9]==locations_summarized[i,8],1:7],2,mean)
}

locations_summarized<-data.frame(locations_summarized)
for(i in 1:7)
{
  locations_summarized[,i]<-as.numeric(as.character(locations_summarized[,i]))
}
colnames(locations_summarized)<-c("lon","lat","horizontal","linear","concave","convex","curved2","LTER")

location_transform<-usmap_transform(locations_summarized[,1:2])

lon2<-c()
lat2<-c()
 for(i in 1:nrow(locations_summarized))
     {
         lon2<-c(lon2,location_transform[location_transform[,1]==locations_summarized[i,1],3])
         lat2<-c(lat2,location_transform[location_transform[,2]==locations_summarized[i,2],4])
 }
locations_summarized<-cbind(locations_summarized,lon2,lat2)


# FIGURE 1 
mapplot4 <- plot_usmap(color = "white", fill = "gray50") + 
  geom_scatterpie(data=locations_summarized,aes(x=lon2, y=lat2), cols = colnames(locations_summarized[,c(4:7)]),pie_scale=2.5) +
  scale_fill_manual(breaks = c("linear", "concave","convex", "curved2"),labels = c("linear", "concave","convex", "curved2"),values = c("linear" = "green","concave" = "darkblue","convex"="lightblue","curved2" = "purple"))+
    coord_fixed() +
    theme_bw()

mapplot4
```

Plots to add to the map:
```{r species example, echo=FALSE}
par(mfrow=c(2,2))
makeTransparent<-function(someColor, alpha=100)
{
  newColor<-col2rgb(someColor)
  apply(newColor, 2, function(curcoldata){rgb(red=curcoldata[1], green=curcoldata[2],
    blue=curcoldata[3],alpha=alpha, maxColorValue=255)})
}

#summarize gam shape by the majority. Two species had a non-clear winner, so we used parsimony and pick the simplest shape. 
gam_majority2<-c()
 for(i in 1:ncol(Type))
     {
         if(summary(factor(Type[,i],levels=c(0,1,2,3)))[1]>50)
             {gam_majority2<-c(gam_majority2,0)}
         if(summary(factor(Type[,i],levels=c(0,1,2,3)))[1]<51)
             {gam_majority2<-c(gam_majority2,1)}
         }

for(i in 1:length(gam_majority2))
     {
         if(gam_majority2[i]==1)
             {
                 gam_majority2[i]<-c(1,2,3)[summary(factor(Type[,i],levels=c(0,1,2,3)))[2:4]==max(summary(factor(Type[,i],levels=c(0,1,2,3)))[2:4])][1]
                 }
}

for(i in 1:length(gam_majority2))
     {
         if(gam_majority2[i]==2)
             {
           if(mean(V[Type[,i]==2,i])<0)
           {gam_majority2[i]<--2}
         }
}

horizontalExample<-as.numeric(K[gam_majority2==0])
linearExample<-as.numeric(K[gam_majority2==1])
sigmoidExample<-as.numeric(K[gam_majority2==3])
positiveExample<-as.numeric(K[gam_majority2==2])
negativeExample<-as.numeric(K[gam_majority2==-2])

  gam_color<-c(NA,makeTransparent("green",90),makeTransparent("blue",90),makeTransparent("purple",90))
    
    N_draws <- 100
  
    #horizontalExample
     V_sum<-RESULTS[[c(1:length(RESULTS))[locations$ID==horizontalExample[547]]]]$V_summary
 V_sum<-data.frame(V_sum)
 for(j in 2:5)
 {
   V_sum[,j]<-as.numeric(as.character(V_sum[,j]))
 }
s<-c(1:nrow(V_sum))[V_sum[,2]==max(V_sum[,2])][2]

r_clim<-RESULTS[[c(1:length(RESULTS))[locations$ID==horizontalExample[547]]]]$r_clim
spp_names<-RESULTS[[c(1:length(RESULTS))[locations$ID==horizontalExample[547]]]]$spp_names
r_posterior<-RESULTS[[c(1:length(RESULTS))[locations$ID==horizontalExample[547]]]]$r_posterior
gt<-RESULTS[[c(1:length(RESULTS))[locations$ID==horizontalExample[547]]]]$gam_type
  
  gam_fit <- gam(r ~ s(SPEI,k=length(unique(r_clim$SPEI[r_clim$species==spp_names[s]]))), data=subset(r_clim,species==spp_names[s]))
  SPEI.seq<-data.frame(SPEI=rep(seq(min(r_clim$SPEI),max(r_clim$SPEI),length=100)))
  preds_mean<-predict(gam_fit, type="terms", newdata=SPEI.seq,se.fit=TRUE)
  plot(r_clim$SPEI[r_clim$species==spp_names[s]],r_clim$r[r_clim$species==spp_names[s]],
       cex.lab=2,xlab="SPEI",ylab="r",type="n",ylim=c(min(r_clim$r[r_clim$species==spp_names[s]]-1),max(r_clim$r[r_clim$species==spp_names[s]]+1)))
  title(main = paste("Species: ",spp_names[s]),adj=0)

  r_clim_species<-r_clim[r_clim$species==spp_names[s],]
  
  for(i in 1:N_draws)
    {
    
    df_i <- data.frame(r_posterior[i,,])  
    colnames(df_i) <- spp_names
    df_i$year <- unique(r_clim$year)
    r_clim_i <- full_join(
      df_i %>% 
      gather(key="species",value="r",1:length(spp_names)) %>% 
      mutate(species=as.factor(species)) %>% 
      #filter(species %in% sig_spp),
      filter(species==spp_names[s]),
       tibble(SPEI = r_clim_species$SPEI,
                               year = unique(r_clim_species$year)),
      by="year")%>% 
      filter(!is.na(r))
    
  gam_fit_i <- gam(r ~ s(SPEI,k=length(unique(r_clim_i$SPEI[r_clim_i$species==spp_names[s]]))), data=subset(r_clim_i,species==spp_names[s]))
  
  preds<-predict(gam_fit_i, type="terms", newdata=SPEI.seq,se.fit=TRUE)
  
  lines(SPEI.seq$SPEI,preds$fit,type="l",col=gam_color[gt[i,s]+1])
  points(r_clim_i$SPEI[r_clim_i$species==spp_names[s]],
         r_clim_i$r[r_clim_i$species==spp_names[s]],pch=".",cex=3,col=gam_color[gt[i,s]+1])

  }
  
  lines(SPEI.seq$SPEI,preds_mean$fit,type="l",lwd=4)
  points(r_clim$SPEI[r_clim$species==spp_names[s]],r_clim$r[r_clim$species==spp_names[s]],
       pch=16,cex=2)

    
    #linearExample 
 V_sum<-RESULTS[[c(1:length(RESULTS))[locations$ID==linearExample[7]]]]$V_summary
 V_sum<-data.frame(V_sum)
 for(j in 2:5)
 {
   V_sum[,j]<-as.numeric(as.character(V_sum[,j]))
 }
s<-15

r_clim<-RESULTS[[c(1:length(RESULTS))[locations$ID==linearExample[7]]]]$r_clim
spp_names<-RESULTS[[c(1:length(RESULTS))[locations$ID==linearExample[7]]]]$spp_names
r_posterior<-RESULTS[[c(1:length(RESULTS))[locations$ID==linearExample[7]]]]$r_posterior
gt<-RESULTS[[c(1:length(RESULTS))[locations$ID==linearExample[7]]]]$gam_type
  
  gam_fit <- gam(r ~ s(SPEI,k=length(unique(r_clim$SPEI[r_clim$species==spp_names[s]]))), data=subset(r_clim,species==spp_names[s]))
  SPEI.seq<-data.frame(SPEI=rep(seq(min(r_clim$SPEI),max(r_clim$SPEI),length=100)))
  preds_mean<-predict(gam_fit, type="terms", newdata=SPEI.seq,se.fit=TRUE)
  plot(r_clim$SPEI[r_clim$species==spp_names[s]],r_clim$r[r_clim$species==spp_names[s]],
       cex.lab=2,xlab="SPEI",ylab="r",type="n",ylim=c(min(r_clim$r[r_clim$species==spp_names[s]]-1),max(r_clim$r[r_clim$species==spp_names[s]]+1)))
  title(main = paste("Species: ",spp_names[s]),adj=0)

  r_clim_species<-r_clim[r_clim$species==spp_names[s],]
  
  for(i in 1:N_draws)
    {
    
    df_i <- data.frame(r_posterior[i,,]) 
    colnames(df_i) <- spp_names
    df_i$year <- unique(r_clim$year)
    r_clim_i <- full_join(
      df_i %>% 
      gather(key="species",value="r",1:length(spp_names)) %>% 
      mutate(species=as.factor(species)) %>% 
      #filter(species %in% sig_spp),
      filter(species==spp_names[s]),
       tibble(SPEI = r_clim_species$SPEI,
                               year = unique(r_clim_species$year)),
      by="year")%>% 
      filter(!is.na(r))
    
  gam_fit_i <- gam(r ~ s(SPEI,k=length(unique(r_clim_i$SPEI[r_clim_i$species==spp_names[s]]))), data=subset(r_clim_i,species==spp_names[s]))
  
  preds<-predict(gam_fit_i, type="terms", newdata=SPEI.seq,se.fit=TRUE)
  
  lines(SPEI.seq$SPEI,preds$fit,type="l",col=gam_color[gt[i,s]+1])
  points(r_clim_i$SPEI[r_clim_i$species==spp_names[s]],
         r_clim_i$r[r_clim_i$species==spp_names[s]],pch=".",cex=3,col=gam_color[gt[i,s]+1])

  }
  
  lines(SPEI.seq$SPEI,preds_mean$fit,type="l",lwd=4)
  points(r_clim$SPEI[r_clim$species==spp_names[s]],r_clim$r[r_clim$species==spp_names[s]],
       pch=16,cex=2)

#positiveExample  
  
  gam_color_pos<-gam_color
  gam_color_pos[3]<-"lightblue"
  
   V_sum<-RESULTS[[c(1:length(RESULTS))[locations$ID==positiveExample[8]]]]$V_summary
    V_sum<-data.frame(V_sum)
 for(j in 2:5)
 {
   V_sum[,j]<-as.numeric(as.character(V_sum[,j]))
 }
s<-41
r_clim<-RESULTS[[c(1:length(RESULTS))[locations$ID==positiveExample[8]]]]$r_clim
spp_names<-RESULTS[[c(1:length(RESULTS))[locations$ID==positiveExample[8]]]]$spp_names
r_posterior<-RESULTS[[c(1:length(RESULTS))[locations$ID==positiveExample[8]]]]$r_posterior
gt<-RESULTS[[c(1:length(RESULTS))[locations$ID==positiveExample[8]]]]$gam_type
  
  gam_fit <- gam(r ~ s(SPEI,k=length(unique(r_clim$SPEI[r_clim$species==spp_names[s]]))), data=subset(r_clim,species==spp_names[s]))
  SPEI.seq<-data.frame(SPEI=rep(seq(min(r_clim$SPEI),max(r_clim$SPEI),length=100)))
  preds_mean<-predict(gam_fit, type="terms", newdata=SPEI.seq,se.fit=TRUE)
  plot(r_clim$SPEI[r_clim$species==spp_names[s]],r_clim$r[r_clim$species==spp_names[s]],
       cex.lab=2,xlab="SPEI",ylab="r",type="n",ylim=c(min(r_clim$r[r_clim$species==spp_names[s]]-1),max(r_clim$r[r_clim$species==spp_names[s]]+1)))
  title(main = paste("Species: ",spp_names[s]),adj=0)

    r_clim_species<-r_clim[r_clim$species==spp_names[s],]
  
  for(i in 1:N_draws)
    {
    
    df_i <- data.frame(r_posterior[i,,])  
    colnames(df_i) <- spp_names
    df_i$year <- unique(r_clim$year)
    r_clim_i <- full_join(
      df_i %>% 
      gather(key="species",value="r",1:length(spp_names)) %>% 
      mutate(species=as.factor(species)) %>% 
      #filter(species %in% sig_spp),
      filter(species==spp_names[s]),
      tibble(SPEI = r_clim_species$SPEI,
                               year = unique(r_clim$year)),
      by="year")%>% 
      filter(!is.na(r))
    
  gam_fit_i <- gam(r ~ s(SPEI,k=length(unique(r_clim_i$SPEI[r_clim_i$species==spp_names[s]]))), data=subset(r_clim_i,species==spp_names[s]))
  
  preds<-predict(gam_fit_i, type="terms", newdata=SPEI.seq,se.fit=TRUE)
  
  lines(SPEI.seq$SPEI,preds$fit,type="l",col=gam_color_pos[gt[i,s]+1])
  points(r_clim_i$SPEI[r_clim_i$species==spp_names[s]],
         r_clim_i$r[r_clim_i$species==spp_names[s]],pch=".",cex=3,col=gam_color_pos[gt[i,s]+1])

  }
  
  lines(SPEI.seq$SPEI,preds_mean$fit,type="l",lwd=4)
  points(r_clim$SPEI[r_clim$species==spp_names[s]],r_clim$r[r_clim$species==spp_names[s]],
       pch=16,cex=2)

  
  #negativeExample  

    gam_color_neg<-gam_color
  gam_color_neg[3]<-"darkblue"  
  
   V_sum<-RESULTS[[c(1:length(RESULTS))[locations$ID==negativeExample[6]]]]$V_summary
    V_sum<-data.frame(V_sum)
 for(j in 2:5)
 {
   V_sum[,j]<-as.numeric(as.character(V_sum[,j]))
 }
s<-c(1:nrow(V_sum))[V_sum[,4]==max(V_sum[,4])]
r_clim<-RESULTS[[c(1:length(RESULTS))[locations$ID==negativeExample[6]]]]$r_clim
spp_names<-RESULTS[[c(1:length(RESULTS))[locations$ID==negativeExample[6]]]]$spp_names
r_posterior<-RESULTS[[c(1:length(RESULTS))[locations$ID==negativeExample[6]]]]$r_posterior
gt<-RESULTS[[c(1:length(RESULTS))[locations$ID==negativeExample[6]]]]$gam_type
  
  gam_fit <- gam(r ~ s(SPEI,k=length(unique(r_clim$SPEI[r_clim$species==spp_names[s]]))), data=subset(r_clim,species==spp_names[s]))
  SPEI.seq<-data.frame(SPEI=rep(seq(min(r_clim$SPEI),max(r_clim$SPEI),length=100)))
  preds_mean<-predict(gam_fit, type="terms", newdata=SPEI.seq,se.fit=TRUE)
  plot(r_clim$SPEI[r_clim$species==spp_names[s]],r_clim$r[r_clim$species==spp_names[s]],
       cex.lab=2,xlab="SPEI",ylab="r",type="n",ylim=c(min(r_clim$r[r_clim$species==spp_names[s]]-1),max(r_clim$r[r_clim$species==spp_names[s]]+1)))
  title(main = paste("Species: ",spp_names[s]),adj=0)

  r_clim_species<-r_clim[r_clim$species==spp_names[s],]
  
  for(i in 1:N_draws)
    {
    
    df_i <- data.frame(r_posterior[i,,])  
    colnames(df_i) <- spp_names
    df_i$year <- unique(r_clim$year)
    r_clim_i <- full_join(
      df_i %>% 
      gather(key="species",value="r",1:length(spp_names)) %>% 
      mutate(species=as.factor(species)) %>% 
      #filter(species %in% sig_spp),
      filter(species==spp_names[s]),
      tibble(SPEI = r_clim_species$SPEI,
                               year = unique(r_clim$year)),
      by="year")%>% 
      filter(!is.na(r))
    
  gam_fit_i <- gam(r ~ s(SPEI,k=length(unique(r_clim_i$SPEI[r_clim_i$species==spp_names[s]]))), data=subset(r_clim_i,species==spp_names[s]))
  
  preds<-predict(gam_fit_i, type="terms", newdata=SPEI.seq,se.fit=TRUE)
  
  lines(SPEI.seq$SPEI,preds$fit,type="l",col=gam_color_neg[gt[i,s]+1])
  points(r_clim_i$SPEI[r_clim_i$species==spp_names[s]],
         r_clim_i$r[r_clim_i$species==spp_names[s]],pch=".",cex=3,col=gam_color_neg[gt[i,s]+1])

  }
  
  lines(SPEI.seq$SPEI,preds_mean$fit,type="l",lwd=4)
  points(r_clim$SPEI[r_clim$species==spp_names[s]],r_clim$r[r_clim$species==spp_names[s]],
       pch=16,cex=2)

  #sigmoidExample  
   V_sum<-RESULTS[[c(1:29)[locations$ID==sigmoidExample[2]]]]$V_summary
    V_sum<-data.frame(V_sum)
 for(j in 2:5)
 {
   V_sum[,j]<-as.numeric(as.character(V_sum[,j]))
 }
s<-c(1:nrow(V_sum))[V_sum[,5]==max(V_sum[,5])]
r_clim<-RESULTS[[c(1:29)[locations$ID==sigmoidExample[2]]]]$r_clim
spp_names<-RESULTS[[c(1:29)[locations$ID==sigmoidExample[2]]]]$spp_names
r_posterior<-RESULTS[[c(1:29)[locations$ID==sigmoidExample[2]]]]$r_posterior
gt<-RESULTS[[c(1:29)[locations$ID==sigmoidExample[2]]]]$gam_type
  
  gam_fit <- gam(r ~ s(SPEI,k=length(unique(r_clim$SPEI[r_clim$species==spp_names[s]]))), data=subset(r_clim,species==spp_names[s]))
  SPEI.seq<-data.frame(SPEI=rep(seq(min(r_clim$SPEI),max(r_clim$SPEI),length=100)))
  preds_mean<-predict(gam_fit, type="terms", newdata=SPEI.seq,se.fit=TRUE)
  plot(r_clim$SPEI[r_clim$species==spp_names[s]],r_clim$r[r_clim$species==spp_names[s]],
       cex.lab=2,xlab="SPEI",ylab="r",type="n",ylim=c(min(r_clim$r[r_clim$species==spp_names[s]]-1),max(r_clim$r[r_clim$species==spp_names[s]]+1)))
  title(main = paste("Species: ",spp_names[s]),adj=0)

  r_clim_species<-r_clim[r_clim$species==spp_names[s],]
  
  for(i in 1:N_draws)
    {
    
    df_i <- data.frame(r_posterior[i,,])  
    colnames(df_i) <- spp_names
    df_i$year <- unique(r_clim$year)
    r_clim_i <- full_join(
      df_i %>% 
      gather(key="species",value="r",1:length(spp_names)) %>% 
      mutate(species=as.factor(species)) %>% 
      #filter(species %in% sig_spp),
      filter(species==spp_names[s]),
      tibble(SPEI = r_clim_species$SPEI,
                               year = unique(r_clim$year)),
      by="year")%>% 
      filter(!is.na(r))
    
  gam_fit_i <- gam(r ~ s(SPEI,k=length(unique(r_clim_i$SPEI[r_clim_i$species==spp_names[s]]))), data=subset(r_clim_i,species==spp_names[s]))
  
  preds<-predict(gam_fit_i, type="terms", newdata=SPEI.seq,se.fit=TRUE)
  
  lines(SPEI.seq$SPEI,preds$fit,type="l",col=gam_color[gt[i,s]+1])
  points(r_clim_i$SPEI[r_clim_i$species==spp_names[s]],
         r_clim_i$r[r_clim_i$species==spp_names[s]],pch=".",cex=3,col=gam_color[gt[i,s]+1])

  }
  
  lines(SPEI.seq$SPEI,preds_mean$fit,type="l",lwd=4)
  points(r_clim$SPEI[r_clim$species==spp_names[s]],r_clim$r[r_clim$species==spp_names[s]],
       pch=16,cex=2)
  
```

## Results through species traits - Figure 2

```{r species traits, echo=FALSE}
setwd("/path")
traits<-read.csv("traits2.csv",header=TRUE)
traits$Species.code[1801]<-"NA"
traits$Species.code[1519]<-"NA"
traits$Species.code[2127]<-"NA"
traits<-traits[traits$proj_metadata_key%in%ID,]
sp<-c()
for(i in 1:length(RESULTS))
{
  sp<-c(sp,RESULTS[[i]]$spp_names)
}  

print(paste("across studies, we obtained data for",length(sp),"species, including",length(unique(sp)),"uniques ones",sep=" "))

traits<-traits[traits$Species.code%in%sp,]

traits<-distinct(traits)

test_combo<-paste(traits$proj_metadata_key,traits$Species.code)
test_data<-paste(K,sp)

traits<-cbind(test_combo,traits)
traits<-traits[traits$test_combo%in%test_data,]

traits_test<-traits[,c(4,6,8,11,12)]
traits_test<-distinct(traits_test)

print(summary(as.factor(traits_test$Kingdom)))
print(summary(as.factor(traits_test$Ecosystem.Type)))
print(summary(as.factor(traits_test$Class)))
print(summary(as.factor(traits_test$Growth.form)))

V<-RESULTS[[1]]$V
type<-RESULTS[[1]]$gam_type

for(i in 2:length(RESULTS))
{
 V<-cbind(V,RESULTS[[i]]$V)
type<-cbind(type,RESULTS[[i]]$gam_type)
}

Kingdom<-rep(NA,length(sp))
  Ecosystem<-rep(NA,length(sp))
      Growth<-rep(NA,length(sp))
        Class<-rep(NA,length(sp))
for(i in 1:length(sp))
{
  if(length(traits$Kingdom[traits$Species.code==sp[i]])!=0)
  {
  Kingdom[i]<-traits$Kingdom[traits$Species.code==sp[i]]
  }
  if(length(traits$Ecosystem.Type[traits$Species.code==sp[i]])!=0)
  {
  Ecosystem[i]<-traits$Ecosystem.Type[traits$Species.code==sp[i]]
  }
    if(length(traits$Growth.form[traits$Species.code==sp[i]])!=0)
  {
  Growth[i]<-traits$Growth.form[traits$Species.code==sp[i]]
    }
      if(length(traits$Class[traits$Species.code==sp[i]])!=0)
  {
  Class[i]<-traits$Class[traits$Species.code==sp[i]]
    }
}
  
VnoNA<-V[,is.na(Kingdom)==FALSE]
typenoNA<-type[,is.na(Kingdom)==FALSE]
boxplot(apply(VnoNA,2,mean)~as.factor(Kingdom[is.na(Kingdom)==FALSE]),ylab="V mean",xlab="",col=c("Brown","Red","Blue","Darkgreen"))

Vkingdom<-apply(VnoNA,2,mean)
by(Vkingdom,as.factor(Kingdom[is.na(Kingdom)==FALSE]),mean)
by(Vkingdom,as.factor(Kingdom[is.na(Kingdom)==FALSE]),quantile,c(0.025,0.975))

pass<-typenoNA[,Kingdom[is.na(Kingdom)==FALSE]=="Plantae"]
tt<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Kingdom[is.na(Kingdom)==FALSE]=="Plantae"]
np<-pass_V[1,pass[1,]==2]
np<-c(length(np[np<0])/length(np),length(np[np>0])/length(np))
 for(i in 2:100)
     {
         tt<-rbind(tt,summary(factor(pass[i,],levels=c(0,1,2,3))))
         np_pass<-pass_V[i,pass[i,]==2]
         np<-rbind(np,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
  tt<-tt/summary(as.factor(Kingdom))[4]   
  np<-np*tt[,3]
  
pass<-typenoNA[,Kingdom[is.na(Kingdom)==FALSE]=="Animalia"]
tt2<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Kingdom[is.na(Kingdom)==FALSE]=="Animalia"]
np2<-pass_V[1,pass[1,]==2]
np2<-c(length(np2[np2<0])/length(np2),length(np2[np2>0])/length(np2))
 for(i in 2:100)
     {
         tt2<-rbind(tt2,summary(factor(pass[i,],levels=c(0,1,2,3))))
                  np_pass<-pass_V[i,pass[i,]==2]
         np2<-rbind(np2,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
  tt2<-tt2/summary(as.factor(Kingdom))[1]  
    np2<-np2*tt2[,3]
    
pass<-typenoNA[,Kingdom[is.na(Kingdom)==FALSE]=="Bacteria"]
tt3<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Kingdom[is.na(Kingdom)==FALSE]=="Bacteria"]
np3<-pass_V[1,pass[1,]==2]
np3<-c(length(np3[np3<0])/length(np3),length(np3[np3>0])/length(np3))
 for(i in 2:100)
     {
         tt3<-rbind(tt3,summary(factor(pass[i,],levels=c(0,1,2,3))))
                           np_pass<-pass_V[i,pass[i,]==2]
         np3<-rbind(np3,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
tt3<-tt3/summary(as.factor(Kingdom))[2]  
 np3<-np3*tt3[,3]
 
pass<-typenoNA[,Kingdom[is.na(Kingdom)==FALSE]=="Chromista"]
tt4<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Kingdom[is.na(Kingdom)==FALSE]=="Chromista"]
np4<-pass_V[1,pass[1,]==2]
np4<-c(length(np4[np4<0])/length(np4),length(np4[np4>0])/length(np4))
 for(i in 2:100)
     {
         tt4<-rbind(tt4,summary(factor(pass[i,],levels=c(0,1,2,3))))
                           np_pass<-pass_V[i,pass[i,]==2]
         np4<-rbind(np4,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
tt4<-tt4/summary(as.factor(Kingdom))[3]  
np4<-np4*tt4[,3]

for(i in 1:100)
{
  if(is.na(np4[i,1])==TRUE)
  {np4[i,1]<-0}
   if(is.na(np3[i,1])==TRUE)
   {np3[i,1]<-0}
   if(is.na(np2[i,1])==TRUE)
   {np2[i,1]<-0}
   if(is.na(np[i,1])==TRUE)
  {np[i,1]<-0}

    if(is.na(np4[i,2])==TRUE)
  {np4[i,2]<-0}
   if(is.na(np3[i,2])==TRUE)
   {np3[i,2]<-0}
   if(is.na(np2[i,2])==TRUE)
   {np2[i,2]<-0}
   if(is.na(np[i,2])==TRUE)
  {np[i,2]<-0}
}

TT<-cbind(tt2[,1],tt3[,1],tt4[,1],tt[,1],rep(NA,100),tt2[,2],tt3[,2],tt4[,2],tt[,2],rep(NA,100),np2[,1],np3[,1],np4[,1],np[,1],rep(NA,100),np2[,2],np3[,2],np4[,2],np[,2],rep(NA,100),tt2[,4],tt3[,4],tt4[,4],tt[,4])

x<-barplot(apply(TT,2,median),ylim=c(0,1),col=c("Brown","Red","Blue","Darkgreen","white","Brown","Red","Blue","Darkgreen","white","Brown","Red","Blue","Darkgreen","white","Brown","Red","Blue","Darkgreen","white","Brown","Red","Blue","Darkgreen"),space=0)
segments(x[-c(5,10,15,20)],apply(TT[,-c(5,10,15,20)],2,quantile,0.025),x[-c(5,10,15,20)],apply(TT[,-c(5,10,15,20)],2,quantile,0.975))
axis(1,at=x[c(2,7,12,17,22),1],labels=c("Flat","Linear","Concave","Convex","Curved2"))
#legend(locator(1),col=c("Brown","Red","Blue","Darkgreen"),legend=c("Animalia","Bacteria","Chromista","Plantae"),pch=20,cex=1.25)

#appendix: Flat
x<-barplot(apply(TT[,1:4],2,median),ylim=c(0,1),col=c("Brown","Red","Blue","Darkgreen"),space=0)
segments(x,apply(TT[,1:4],2,quantile,0.025),x,apply(TT[,1:4],2,quantile,0.975))
axis(1,at=2,labels=c("Flat"))

# add a few tests: Plant versus animals
t.test(TT[,1],TT[,4])
t.test(TT[,6],TT[,9])
t.test(TT[,11],TT[,14])
t.test(TT[,16],TT[,19])
t.test(TT[,21],TT[,24])

Ecosystem[Ecosystem==""]<-NA
Ecosystem[Ecosystem=="Marine"]<-"Aquatic"
VnoNA<-V[,is.na(Ecosystem)==FALSE]
typenoNA<-type[,is.na(Ecosystem)==FALSE]
boxplot(apply(VnoNA,2,mean)~as.factor(Ecosystem[is.na(Ecosystem)==FALSE]),ylab="V mean",xlab="",col=c("Darkblue","Darkgreen","Brown"))

pass<-typenoNA[,Ecosystem[is.na(Ecosystem)==FALSE]%in%c("Aquatic","Marine")]
tt<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Ecosystem[is.na(Ecosystem)==FALSE]%in%c("Aquatic","Marine")]
np<-pass_V[1,pass[1,]==2]
np<-c(length(np[np<0])/length(np),length(np[np>0])/length(np))
 for(i in 2:100)
     {
         tt<-rbind(tt,summary(factor(pass[i,],levels=c(0,1,2,3))))
          np_pass<-pass_V[i,pass[i,]==2]
         np<-rbind(np,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
  tt<-tt/(summary(as.factor(Ecosystem))[1]) 
    np<-np*tt[,3]
  
pass<-typenoNA[,Ecosystem[is.na(Ecosystem)==FALSE]=="Semiaquatic"]
tt2<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Ecosystem[is.na(Ecosystem)==FALSE]=="Semiaquatic"]
np2<-pass_V[1,pass[1,]==2]
np2<-c(length(np2[np2<0])/length(np2),length(np2[np2>0])/length(np2))
 for(i in 2:100)
     {
         tt2<-rbind(tt2,summary(factor(pass[i,],levels=c(0,1,2,3))))
                   np_pass<-pass_V[i,pass[i,]==2]
         np2<-rbind(np2,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
  tt2<-tt2/summary(as.factor(Ecosystem))[2] 
  np2<-np2*tt2[,3]
  
pass<-typenoNA[,Ecosystem[is.na(Ecosystem)==FALSE]=="Terrestrial"]
tt3<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Ecosystem[is.na(Ecosystem)==FALSE]=="Terrestrial"]
np3<-pass_V[1,pass[1,]==2]
np3<-c(length(np3[np3<0])/length(np3),length(np3[np3>0])/length(np3))
 for(i in 2:100)
     {
         tt3<-rbind(tt3,summary(factor(pass[i,],levels=c(0,1,2,3))))
            np_pass<-pass_V[i,pass[i,]==2]
         np3<-rbind(np3,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
tt3<-tt3/summary(as.factor(Ecosystem))[3]  
np3<-np3*tt3[,3]

TT<-cbind(tt[,1],tt2[,1],tt3[,1],rep(NA,100),tt[,2],tt2[,2],tt3[,2],rep(NA,100),np[,1],np2[,1],np3[,1],rep(NA,100),np[,2],np2[,2],np3[,2],rep(NA,100),tt[,4],tt2[,4],tt3[,4])

x<-barplot(apply(TT,2,median),ylim=c(0,1),col=c("Darkblue","DarkGreen","Brown","white","Darkblue","Darkgreen","Brown","white","Darkblue","Darkgreen","Brown","white","Darkblue","Darkgreen","Brown","white","Darkblue","Darkgreen","Brown"),space=0)
segments(x[-c(4,8,12,16)],apply(TT[,-c(4,8,12,16)],2,quantile,0.025),x[-c(4,8,12,16)],apply(TT[,-c(4,8,12,16)],2,quantile,0.975))
axis(1,at=x[c(2,6,10,14,18),1],labels=c("Flat","Linear","Concave","Convex","Curved2"))
#legend(locator(1),col=c("Darkblue","DarkGreen","Brown"),legend=c("Aquatic","Seimaquatic","Terrestrial"),pch=20,cex=1.25)

#appendix: Flat
x<-barplot(apply(TT[,1:3],2,median),ylim=c(0,1),col=c("Darkblue","DarkGreen","Brown"),space=0)
segments(x,apply(TT[,1:3],2,quantile,0.025),x,apply(TT[,1:3],2,quantile,0.975))
axis(1,at=1.5,labels=c("Flat"))


# add a few tests
t.test(TT[,1],TT[,3])
t.test(TT[,5],TT[,7])
t.test(TT[,9],TT[,11])
t.test(TT[,13],TT[,15])
t.test(TT[,17],TT[,19])

#CLASS
Class[Class==""]<-NA
Class[(Kingdom=="Animalia")==FALSE]<-NA
Class[(Class%in%c("Actinopterygii","Aves","Insecta"))==FALSE]<-NA
VnoNA<-V[,is.na(Class)==FALSE]
typenoNA<-type[,is.na(Class)==FALSE]
boxplot(apply(VnoNA,2,mean)~as.factor(Class[is.na(Class)==FALSE]),ylab="V mean",xlab="",col=c("Darkblue","Darkgreen","Brown"))
test<-apply(VnoNA,2,mean)
by(test,Class[is.na(Class)==FALSE],mean)
by(test,Class[is.na(Class)==FALSE],quantile,c(0.025,0.975))

pass<-typenoNA[,Class[is.na(Class)==FALSE]%in%c("Actinopterygii")]
tt<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Class[is.na(Class)==FALSE]%in%c("Actinopterygii")]
np<-pass_V[1,pass[1,]==2]
np<-c(length(np[np<0])/length(np),length(np[np>0])/length(np))
 for(i in 2:100)
     {
         tt<-rbind(tt,summary(factor(pass[i,],levels=c(0,1,2,3))))
          np_pass<-pass_V[i,pass[i,]==2]
         np<-rbind(np,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
  tt<-tt/(summary(as.factor(Class))[1]) 
    np<-np*tt[,3]
  
pass<-typenoNA[,Class[is.na(Class)==FALSE]%in%c("Aves")]
tt2<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Class[is.na(Class)==FALSE]%in%c("Aves")]
np2<-pass_V[1,pass[1,]==2]
np2<-c(length(np2[np2<0])/length(np2),length(np2[np2>0])/length(np2))
 for(i in 2:100)
     {
         tt2<-rbind(tt2,summary(factor(pass[i,],levels=c(0,1,2,3))))
                   np_pass<-pass_V[i,pass[i,]==2]
         np2<-rbind(np2,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
  tt2<-tt2/summary(as.factor(Class))[2] 
  np2<-np2*tt2[,3]
  
pass<-typenoNA[,Class[is.na(Class)==FALSE]%in%c("Insecta")]
tt3<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Class[is.na(Class)==FALSE]%in%c("Insecta")]
np3<-pass_V[1,pass[1,]==2]
np3<-c(length(np3[np3<0])/length(np3),length(np3[np3>0])/length(np3))
 for(i in 2:100)
     {
         tt3<-rbind(tt3,summary(factor(pass[i,],levels=c(0,1,2,3))))
            np_pass<-pass_V[i,pass[i,]==2]
         np3<-rbind(np3,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
tt3<-tt3/summary(as.factor(Class))[3]  
np3<-np3*tt3[,3]

TT<-cbind(tt[,1],tt2[,1],tt3[,1],rep(NA,100),tt[,2],tt2[,2],tt3[,2],rep(NA,100),np[,1],np2[,1],np3[,1],rep(NA,100),np[,2],np2[,2],np3[,2],rep(NA,100),tt[,4],tt2[,4],tt3[,4])

x<-barplot(apply(TT,2,median),ylim=c(0,1),col=c("Darkblue","DarkGreen","Brown","white","Darkblue","Darkgreen","Brown","white","Darkblue","Darkgreen","Brown","white","Darkblue","Darkgreen","Brown","white","Darkblue","Darkgreen","Brown"),space=0)
segments(x[-c(4,8,12,16)],apply(TT[,-c(4,8,12,16)],2,quantile,0.025),x[-c(4,8,12,16)],apply(TT[,-c(4,8,12,16)],2,quantile,0.975))
axis(1,at=x[c(2,6,10,14,18),1],labels=c("Flat","Linear","Concave","Convex","Curved2"))
#legend(locator(1),col=c("Darkblue","DarkGreen","Brown"),legend=c("Actinopterygii","Aves","Insecta"),pch=20,cex=1.25)

#appendix: Flat
x<-barplot(apply(TT[,1:3],2,median),ylim=c(0,1),col=c("Darkblue","DarkGreen","Brown"),space=0)
segments(x,apply(TT[,1:3],2,quantile,0.025),x,apply(TT[,1:3],2,quantile,0.975))
axis(1,at=1.5,labels=c("Flat"))

#add some tests
#fish - insect
t.test(TT[,1],TT[,3])
t.test(TT[,5],TT[,7])
t.test(TT[,9],TT[,11])
t.test(TT[,13],TT[,15])
t.test(TT[,17],TT[,19])

#fish - bird
t.test(TT[,1],TT[,2])
t.test(TT[,5],TT[,6])
t.test(TT[,9],TT[,10])
t.test(TT[,13],TT[,14])
t.test(TT[,17],TT[,18])


# bird - insect
t.test(TT[,3],TT[,2])
t.test(TT[,7],TT[,6])
t.test(TT[,11],TT[,10])
t.test(TT[,15],TT[,14])
t.test(TT[,19],TT[,18])

#GROWTH
Growth[Growth==""]<-NA
Growth[Growth=="Vine"]<-NA
Growth[Growth=="Moss"]<-NA
VnoNA<-V[,is.na(Growth)==FALSE]
typenoNA<-type[,is.na(Growth)==FALSE]
boxplot(apply(VnoNA,2,mean)~as.factor(Growth[is.na(Growth)==FALSE]),ylab="V mean",xlab="",col=c("Darkblue","Lightgreen","Green","Darkgreen","Brown"))

Vgrowth<-apply(VnoNA,2,mean)
by(Vgrowth,as.factor(Growth[is.na(Growth)==FALSE]),mean)
by(Vgrowth,as.factor(Growth[is.na(Growth)==FALSE]),quantile,c(0.025,0.975))

pass<-typenoNA[,Growth[is.na(Growth)==FALSE]=="Algae"]
tt<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Growth[is.na(Growth)==FALSE]=="Algae"]
np<-pass_V[1,pass[1,]==2]
np<-c(length(np[np<0])/length(np),length(np[np>0])/length(np))
 for(i in 2:100)
     {
         tt<-rbind(tt,summary(factor(pass[i,],levels=c(0,1,2,3))))
         np_pass<-pass_V[i,pass[i,]==2]
         np<-rbind(np,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
  tt<-tt/summary(as.factor(Growth))[1]   
  np<-np*tt[,3]
  
pass<-typenoNA[,Growth[is.na(Growth)==FALSE]=="Forb"]
tt2<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Growth[is.na(Growth)==FALSE]=="Forb"]
np2<-pass_V[1,pass[1,]==2]
np2<-c(length(np2[np2<0])/length(np2),length(np2[np2>0])/length(np2))
 for(i in 2:100)
     {
         tt2<-rbind(tt2,summary(factor(pass[i,],levels=c(0,1,2,3))))
                  np_pass<-pass_V[i,pass[i,]==2]
         np2<-rbind(np2,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
  tt2<-tt2/summary(as.factor(Growth))[2]  
    np2<-np2*tt2[,3]
    
pass<-typenoNA[,Growth[is.na(Growth)==FALSE]=="Grass"]
tt3<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Growth[is.na(Growth)==FALSE]=="Grass"]
np3<-pass_V[1,pass[1,]==2]
np3<-c(length(np3[np3<0])/length(np3),length(np3[np3>0])/length(np3))
 for(i in 2:100)
     {
         tt3<-rbind(tt3,summary(factor(pass[i,],levels=c(0,1,2,3))))
                           np_pass<-pass_V[i,pass[i,]==2]
         np3<-rbind(np3,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
tt3<-tt3/summary(as.factor(Growth))[3]  
 np3<-np3*tt3[,3]
 
pass<-typenoNA[,Growth[is.na(Growth)==FALSE]=="Shrub"]
tt4<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Growth[is.na(Growth)==FALSE]=="Shrub"]
np4<-pass_V[1,pass[1,]==2]
np4<-c(length(np4[np4<0])/length(np4),length(np4[np4>0])/length(np4))
 for(i in 2:100)
     {
         tt4<-rbind(tt4,summary(factor(pass[i,],levels=c(0,1,2,3))))
                           np_pass<-pass_V[i,pass[i,]==2]
         np4<-rbind(np4,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
tt4<-tt4/summary(as.factor(Growth))[4]  
np4<-np4*tt4[,3]

pass<-typenoNA[,Growth[is.na(Growth)==FALSE]=="Tree"]
tt5<-summary(factor(pass[1,],levels=c(0,1,2,3)))
pass_V<-VnoNA[,Growth[is.na(Growth)==FALSE]=="Tree"]
np5<-pass_V[1,pass[1,]==2]
np5<-c(length(np5[np5<0])/length(np5),length(np5[np5>0])/length(np5))
 for(i in 2:100)
     {
         tt5<-rbind(tt5,summary(factor(pass[i,],levels=c(0,1,2,3))))
                           np_pass<-pass_V[i,pass[i,]==2]
         np5<-rbind(np5,c(length(np_pass[np_pass<0])/length(np_pass),length(np_pass[np_pass>0])/length(np_pass)))
         }
tt5<-tt5/summary(as.factor(Growth))[5]  
np5<-np5*tt5[,3]

for(i in 1:100)
{
    if(is.na(np5[i,1])==TRUE)
  {np5[i,1]<-0}
  if(is.na(np4[i,1])==TRUE)
  {np4[i,1]<-0}
   if(is.na(np3[i,1])==TRUE)
   {np3[i,1]<-0}
   if(is.na(np2[i,1])==TRUE)
   {np2[i,1]<-0}
   if(is.na(np[i,1])==TRUE)
  {np[i,1]<-0}

    if(is.na(np4[i,2])==TRUE)
    {np4[i,2]<-0}
      if(is.na(np5[i,2])==TRUE)
  {np5[i,2]<-0}
   if(is.na(np3[i,2])==TRUE)
   {np3[i,2]<-0}
   if(is.na(np2[i,2])==TRUE)
   {np2[i,2]<-0}
   if(is.na(np[i,2])==TRUE)
  {np[i,2]<-0}
}

TT<-cbind(tt[,1],tt2[,1],tt3[,1],tt4[,1],tt5[,1],rep(NA,100),tt[,2],tt2[,2],tt3[,2],tt4[,2],tt5[,2],rep(NA,100),np[,1],np2[,1],np3[,1],np4[,1],np5[,1],rep(NA,100),np[,2],np2[,2],np3[,2],np4[,2],np5[,2],rep(NA,100),tt[,4],tt2[,4],tt3[,4],tt4[,4],tt5[,4])

x<-barplot(apply(TT,2,median),ylim=c(0,1),col=c("Darkblue","Lightgreen","Green","Darkgreen","Brown","white","Darkblue","Lightgreen","Green","Darkgreen","Brown","white","Darkblue","Lightgreen","Green","Darkgreen","Brown","white","Darkblue","Lightgreen","Green","Darkgreen","Brown","white","Darkblue","Lightgreen","Green","Darkgreen","Brown"),space=0)
segments(x[-c(6,12,18,24)],apply(TT[,-c(6,12,18,24)],2,quantile,0.025),x[-c(6,12,18,24)],apply(TT[,-c(6,12,18,24)],2,quantile,0.975))
axis(1,at=x[c(3,9,15,21,27),1],labels=c("Flat","Linear","Concave","Convex","Curved2"))
#legend(locator(1),col=c("Darkblue","Lightgreen","Green","Darkgreen","Brown"),legend=c("Algae","Forb","Grass","Shrub","Tree"),pch=20,cex=1.25)

#appendix: Flat
x<-barplot(apply(TT[,1:5],2,median),ylim=c(0,1),col=c("Darkblue","Lightgreen","Green","Darkgreen","Brown"),space=0)
segments(x,apply(TT[,1:5],2,quantile,0.025),x,apply(TT[,1:5],2,quantile,0.975))
axis(1,at=2.5,labels=c("Flat"))


#Add some test
# grass - tree
t.test(TT[,3],TT[,5]) #horizontal
t.test(TT[,9],TT[,11]) #linear
t.test(TT[,15],TT[,17]) #concavs
t.test(TT[,21],TT[,23]) #convex, n.s.
t.test(TT[,27],TT[,29]) #interaction

# grass - shrub
t.test(TT[,3],TT[,4]) #horizontal
t.test(TT[,9],TT[,10]) #linear
t.test(TT[,15],TT[,16]) #concavs
t.test(TT[,21],TT[,22]) #convex
t.test(TT[,27],TT[,28]) #interaction

# forb - tree
t.test(TT[,2],TT[,5]) #horizontal
t.test(TT[,8],TT[,11]) #linear, n.s
t.test(TT[,14],TT[,17]) #concavs, n.s
t.test(TT[,20],TT[,23]) #convex, n.s.
t.test(TT[,26],TT[,29]) #interaction, ns

# forb - shrub
t.test(TT[,2],TT[,4]) #horizontal
t.test(TT[,8],TT[,10]) #linear
t.test(TT[,14],TT[,16]) #concavs
t.test(TT[,20],TT[,22]) #convex
t.test(TT[,26],TT[,28]) #interaction

# forb - grass
t.test(TT[,2],TT[,3]) #horizontal
t.test(TT[,8],TT[,9]) #linear
t.test(TT[,14],TT[,15]) #concavs
t.test(TT[,20],TT[,21]) #convex
t.test(TT[,26],TT[,27]) #interaction

```


## Summarizing the var x mean effects - Figure 3

```{r var x mean , echo=FALSE}
setwd("/Path")
mean_effect<-c()
var_effect<-c()
interaction_effect<-c()
for(i in 1:length(studies))
{
  load(studies[i])
  if(dim(output$mean_var_effect)[1]>1)
  {
  mean_effect<-cbind(mean_effect,t(output$mean_var_effect[,,1]))
  var_effect<-cbind(var_effect,t(output$mean_var_effect[,,2]))
  interaction_effect<-cbind(interaction_effect,t(output$mean_var_effect[,,3]))
  }
    if(dim(output$mean_var_effect)[1]==1)
  {
  mean_effect<-cbind(mean_effect,output$mean_var_effect[,,1])
  var_effect<-cbind(var_effect,output$mean_var_effect[,,2])
  interaction_effect<-cbind(interaction_effect,output$mean_var_effect[,,3])
  }
}

mean_effect_save<-mean_effect
var_effect_save<-var_effect
interaction_effect_save<-interaction_effect

#Only focus on significant gams
for( i in 1:ncol(mean_effect))
{
  mean_effect[Type[,i]==0,i]<-0
var_effect[Type[,i]==0,i]<-0
interaction_effect[Type[,i]==0,i]<-0
}


effects<-cbind(apply(mean_effect,2,mean,na.rm=TRUE),apply(var_effect,2,mean,na.rm=TRUE),apply(interaction_effect,2,mean,na.rm=TRUE))
colnames(effects)<-c("Mean","Var","Interaction")
b<-boxplot(effects,plot=FALSE)
b$stats[2,]<-apply(effects,2,quantile,0.025,na.rm=TRUE)
b$stats[4,]<-apply(effects,2,quantile,0.975,na.rm=TRUE)
b$stats[1,]<-apply(effects,2,min,na.rm=TRUE)
b$stats[5,]<-apply(effects,2,max,na.rm=TRUE)
b1<-b$out[b$group==1]
b1<-b1[b1>=b$stats[4,1]]
b2<-b$out[b$group==2]
b2<-b2[b2<=b$stats[2,2]]
b3<-b$out[b$group==2]
b3<-b3[b3>=b$stats[4,3]]
b$group<-c(rep(1,length(b1)),rep(2,length(b2)),rep(3,length(b3)))
b$out<-c(b1,b2,b3)
bxp(b,boxfill=c("green","Blue","purple"))
```


## Forcast - Figure 4

Let's think about how to summarize temporally the historical results of changing V by separating sigmoid, concave, convex along with the climate forecast.

```{r historical time series , echo=FALSE}

## separating by type (sigmoid, concave, convex)
#sigmoid
 V_past<-RESULTS[[1]]$V_ts[,1:94,1:100]
 for( i in 2:length(studies))
 {
    V_past<-abind(V_past,RESULTS[[i]]$V_ts[,1:94,1:100],along=1)
 }
 
V_pass<-V_past[gam_majority2==3,,] 
V_pass_sig<-V[,gam_majority2==3]


Proportions_positive<-c()
Proportions_negative<-c()
Difference_sigmoid<-c()
value_sigmoid<-c()
for(i in 1:100)
{
  Proportions_positive_pass<-c()
  Proportions_negative_pass<-c()
  Difference_sigmoid_pass<-c()
  value_sigmoid_pass<-c()
  for(k in 1:94)
  {
    Proportions_positive_pass<-c(Proportions_positive_pass,length(V_pass[V_pass[,k,i]>0,k,i])/nrow(V_pass))
    Proportions_negative_pass<-c(Proportions_negative_pass,length(V_pass[V_pass[,k,i]<0,k,i])/nrow(V_pass))
    Difference_sigmoid_pass<-c(Difference_sigmoid_pass,mean(V_pass[,k,i]-t(V_pass_sig[i,])))
    value_sigmoid_pass<-c(value_sigmoid_pass,mean(V_pass[,k,i]))
  }
  Proportions_positive<-rbind(Proportions_positive,Proportions_positive_pass)
    Proportions_negative<-rbind(Proportions_negative,Proportions_negative_pass)
    Difference_sigmoid<-rbind(Difference_sigmoid,Difference_sigmoid_pass)
    value_sigmoid<-rbind(value_sigmoid,value_sigmoid_pass)
}


tt0<-apply(Proportions_positive,2,median)
tt0b<-apply(Proportions_positive,2,quantile,0.025)
tt0c<-apply(Proportions_positive,2,quantile,0.975)
year<-c(1901:1994)
type<-rep("1positive",94)
tt0<-cbind(year,tt0,tt0b,tt0c,type)
tt0<-data.frame(tt0)
colnames(tt0)<-c("year","median","lower","upper","type")

tt1<-apply(Proportions_negative,2,median)
tt1b<-apply(Proportions_negative,2,quantile,0.025)
tt1c<-apply(Proportions_negative,2,quantile,0.975)
year<-c(1901:1994)
type<-rep("2negative",94)
tt1<-cbind(year,tt1,tt1b,tt1c,type)
tt1<-data.frame(tt1)
colnames(tt1)<-c("year","median","lower","upper","type")

tt0<-rbind(tt0,tt1)
for(i in 1:4)
{
tt0[,i]<-as.numeric(tt0[,i])
}

p0<-ggplot(data=tt0, aes(x=year, y=median, group=type,color=type))  
p0<-p0+geom_ribbon(aes(ymin=lower, ymax=upper,fill=type,color=type), colour=NA, alpha=0.6)+ geom_point() + geom_line()+scale_colour_manual(values = c("#44AA99","#DDCC77")) + scale_fill_manual(values = c("#44AA99","#DDCC77"))

#concave and convex

V_past_neg<-V_past[gam_majority2==-2,,]
V_pass_neg<-V[,gam_majority2==-2]
V_past_neg_diff<-c()
V_past_neg_value<-c()
for(i in 1:100)
{
  V_past_neg_diff_pass<-c()
  V_past_neg_value_pass<-c()
  for(k in 1:94)
  {
    V_past_neg_diff_pass<-c(V_past_neg_diff_pass,mean((V_past_neg[,k,i]-t(V_pass_neg[i,]))))
    V_past_neg_value_pass<-c(V_past_neg_value_pass,mean(V_past_neg[,k,i]))
  }
  V_past_neg_diff<-rbind(V_past_neg_diff,V_past_neg_diff_pass)
  V_past_neg_value<-rbind(V_past_neg_value,V_past_neg_value_pass)
}

tt1_n<-apply(V_past_neg_diff,2,median)
tt1b_n<-apply(V_past_neg_diff,2,quantile,0.025)
tt1c_n<-apply(V_past_neg_diff,2,quantile,0.975)
year<-c(1901:1994)
type<-rep("2negative",94)
tt1_n<-cbind(year,tt1_n,tt1b_n,tt1c_n,type)
tt1_n<-data.frame(tt1_n)
colnames(tt1_n)<-c("year","median","lower","upper","type")

V_past_pos<-V_past[gam_majority2==2,,]
V_pass_pos<-V[,gam_majority2==2]
V_past_pos_diff<-c()
V_past_pos_value<-c()
for(i in 1:100)
{
  V_past_pos_diff_pass<-c()
   V_past_pos_value_pass<-c()
  for(k in 1:94)
  {
    V_past_pos_diff_pass<-c(V_past_pos_diff_pass,mean((V_past_pos[,k,i]-t(V_pass_pos[i,]))))
    V_past_pos_value_pass<-c(V_past_pos_value_pass,mean(V_past_pos[,k,i]))
  }
  V_past_pos_diff<-rbind(V_past_pos_diff,V_past_pos_diff_pass)
  V_past_pos_value<-rbind(V_past_pos_value,V_past_pos_value_pass)
}

tt1_p<-apply(V_past_pos_diff,2,median)
tt1b_p<-apply(V_past_pos_diff,2,quantile,0.025)
tt1c_p<-apply(V_past_pos_diff,2,quantile,0.975)
year<-c(1901:1994)
type<-rep("1positive",94)
tt1_p<-cbind(year,tt1_p,tt1b_p,tt1c_p,type)
tt1_p<-data.frame(tt1_p)
colnames(tt1_p)<-c("year","median","lower","upper","type")

tt1_np<-rbind(tt1_n,tt1_p)

for(i in 1:4)
{
 tt1_np[,i]<-as.numeric(tt1_np[,i]) 
}

p0b<-ggplot(data=tt1_np, aes(x=year, y=median, group=type,color=type))  
p0b<-p0b+geom_ribbon(aes(ymin=lower, ymax=upper,fill=type,color=type), colour=NA, alpha=0.6)+ geom_point() + geom_line()+scale_colour_manual(values = c("#44AA99","#DDCC77")) + scale_fill_manual(values = c("#44AA99","#DDCC77"))

```

Let's forecast the future!
```{r future , echo=FALSE}
# first step get the future climate for the 8 models x 2 climate scenarios and calculate SPEI.

setwd("/Path")
prismfuture<-read.csv("prismfuture.csv",header=TRUE)
V_future_s<-array(data=0,dim=c(1848,912,100))
Mean_Var_future_s<-array(data=NA,dim=c(28,912,100,3))
scenarios<-unique(prismfuture$Scenario)
win_size <- 24
for(w in 1:length(scenarios))
{
  climatefuture <- prismfuture %>% 
     dplyr::select(-X) %>% 
    filter(Scenario == scenarios[w])
S<-0  
for(i in 1:45) #no climate for VCR.
{
climatefuture_pass <- climatefuture %>% 
    filter(ID1 == RESULTS[[i]][1]) %>% 
    group_by(month,Year) %>%  # group by everything other than the value column. 
     dplyr::summarise(tmean=mean(tmean,na.rm=TRUE),
              ppt=(mean(ppt,na.rm=TRUE)))%>%
    ##thornthwaite function does not like NAs
    filter(!is.na(tmean),
           !is.na(ppt))%>% 
    distinct() %>% 
    ##df with years and months for ppt and temp
    arrange_at("month") %>% 
    arrange_at("Year") %>%
    as.data.frame

PET <- thornthwaite(Tave=climatefuture_pass$tmean,RESULTS[[i]]$site_lat)
BAL <- climatefuture_pass$ppt-PET
speiFuture <- spei(BAL, 12)
SPEI = speiFuture$fitted[seq(from=12,to=length(speiFuture$fitted),by=12)]
year<-c(2011:2100)
speiFuture<-data.frame(cbind(SPEI,year))
spp_names<-RESULTS[[i]]$spp_names
r_clim<-RESULTS[[i]]$r_clim

for(s in 1:length(spp_names))
{
 
  for(p in 1:100)
  {
  
    if(length(spp_names)>1)
    {
  r_posterior<-RESULTS[[i]]$r_posterior[p,,s]
    }
        if(length(spp_names)==1)
    {
  r_posterior<-RESULTS[[i]]$r_posterior[p,]
        }
    pass<-c()
    mm<-c()
    vv<-c()
    for(t in 1:(length(year)-win_size))
    {
    window_years <- speiFuture$year[t]:speiFuture$year[t+win_size]
    SPEI_window <- speiFuture %>% filter(year %in% window_years)
    mm<-c(mm,mean(SPEI_window$SPEI))
    vv<-c(vv,var(SPEI_window$SPEI))
    SPEI_window_mean = data.frame(mean(SPEI_window$SPEI));names(SPEI_window_mean)[1]="SPEI"
  data_pass<-subset(r_clim,species==spp_names[s])
  data_pass$r<-r_posterior
  gam_fit<-gam(r ~ s(SPEI,k=length(unique(r_clim$SPEI[r_clim$species==spp_names[s]]))), data=data_pass)
  
  pass<-c(pass,mean(predict(gam_fit, type="terms", newdata=SPEI_window,se.fit=TRUE)$fit) - 
  predict(gam_fit, type="terms", newdata=SPEI_window_mean,se.fit=TRUE)$fit)
    }
  V_future_s[((w-1)*66+1):(w*66),S-length(spp_names)+s,p] <- pass
  Mean_Var_future_s[w,S-length(spp_names)+s,p,1:3]<-rsq.partial(lm(pass~mm*vv))[[3]]
  
  }
}

S<-S+length(spp_names)
}
  
}


#sigmoid

rcp<-scenarios
for(i in 1:28)
      {
 rcp[i]<-substr(rcp[i],nchar(rcp[i])-4,nchar(rcp[i]))
 }
rcp<-rep(rcp,each=66)

V_future_s_3<-V_future_s[,gam_majority2==3,]
V_future_pass<-V_future_s_3
V_pass_sig<-V[,gam_majority2==3]

V_future_s_3_difference<-V_future_s_3
for(k in 1:1848)
{
  V_future_s_3_difference[k,,]<-V_future_s_3[k,,]-t(V_pass_sig[,])
}
V_future_s_3_difference_RCP45<-V_future_s_3_difference[rcp=="RCP45",,]
V_future_s_3_difference_RCP85<-V_future_s_3_difference[rcp=="RCP85",,]
V_future_s_3_value_RCP45<-V_future_s_3[rcp=="RCP45",,]
V_future_s_3_value_RCP85<-V_future_s_3[rcp=="RCP85",,]

Projection_results_n<-matrix(data=NA,nrow=length(scenarios),ncol=66)
Projection_results_p<-matrix(data=NA,nrow=length(scenarios),ncol=66)
Projection_diff_sigmoid<-matrix(data=NA,nrow=length(scenarios),ncol=66)
Projection_sigmoid_value<-matrix(data=NA,nrow=length(scenarios),ncol=66)

for(w in 1:length(scenarios))
{
  V_future_negative<-c()
  V_future_positive<-c()
  V_future_diff<-c()
  V_future_value<-c()
  for(j in ((w-1)*66+1):(w*66))
  {
     pass<-V_future_pass[j,,]
V_future_negative<-c(V_future_negative,length(pass[is.na(pass)==FALSE & pass<0])/length(pass))
V_future_positive<-c(V_future_positive,length(pass[is.na(pass)==FALSE & pass>0])/length(pass))
V_future_diff<-c(V_future_diff,mean(pass-t(V_pass_sig)))
V_future_value<-c(V_future_value,mean(pass))
  }
Projection_results_n[w,]<-V_future_negative
Projection_results_p[w,]<-V_future_positive
Projection_diff_sigmoid[w,]<-V_future_diff
Projection_sigmoid_value[w,]<-V_future_value
}

Projection_results_n_45<-Projection_results_n[c(1,3:7,13,15,17,19:23),]
Projection_results_n_85<-Projection_results_n[c(2,8:12,14,16,18,24:28),]
Projection_results_p_45<-Projection_results_p[c(1,3:7,13,15,17,19:23),]
Projection_results_p_85<-Projection_results_p[c(2,8:12,14,16,18,24:28),]
Projection_diff_sigmoid_45<-Projection_diff_sigmoid[c(1,3:7,13,15,17,19:23),]
Projection_diff_sigmoid_85<-Projection_diff_sigmoid[c(2,8:12,14,16,18,24:28),]
Projection_sigmoid_value_45<-Projection_sigmoid_value[c(1,3:7,13,15,17,19:23),]
Projection_sigmoid_value_85<-Projection_sigmoid_value[c(2,8:12,14,16,18,24:28),]

Projection_results_p_45_m<-apply(Projection_results_p_45,2,mean)
Projection_results_p_45_l<-apply(Projection_results_p_45,2,quantile,0.025)
Projection_results_p_45_h<-apply(Projection_results_p_45,2,quantile,0.975)
tt<-cbind(c(2035:2100),Projection_results_p_45_m,Projection_results_p_45_l,Projection_results_p_45_h,rep("1Positive",66))

tt<-data.frame(tt)
colnames(tt)<-c("year","median","lower","upper","type")
for(i in 1:4)
{
  tt[,i]<-as.numeric(tt[,i])
}

Projection_results_n_45_m<-apply(Projection_results_n_45,2,mean)
Projection_results_n_45_l<-apply(Projection_results_n_45,2,quantile,0.025)
Projection_results_n_45_h<-apply(Projection_results_n_45,2,quantile,0.975)
tt2<-cbind(c(2035:2100),Projection_results_n_45_m,Projection_results_n_45_l,Projection_results_n_45_h,rep("2Negative",66))

tt2<-data.frame(tt2)
colnames(tt2)<-c("year","median","lower","upper","type")
for(i in 1:4)
{
  tt2[,i]<-as.numeric(tt2[,i])
}

tt<-rbind(tt,tt2)

p<-ggplot(data=tt, aes(x=year, y=median, group=type,color=type))  
p<-p+geom_ribbon(aes(ymin=lower, ymax=upper,fill=type), colour=NA, alpha=0.6)+ geom_point() + geom_line()+scale_colour_manual(values = c("#44AA99","#DDCC77")) + scale_fill_manual(values = c("#44AA99","#DDCC77"))

Projection_results_p_85_m<-apply(Projection_results_p_85,2,mean)
Projection_results_p_85_l<-apply(Projection_results_p_85,2,quantile,0.025)
Projection_results_p_85_h<-apply(Projection_results_p_85,2,quantile,0.975)
tt3<-cbind(c(2035:2100),Projection_results_p_85_m,Projection_results_p_85_l,Projection_results_p_85_h,rep("1Positive",66))

tt3<-data.frame(tt3)
colnames(tt3)<-c("year","median","lower","upper","type")
for(i in 1:4)
{
  tt3[,i]<-as.numeric(tt3[,i])
}

Projection_results_n_85_m<-apply(Projection_results_n_85,2,mean)
Projection_results_n_85_l<-apply(Projection_results_n_85,2,quantile,0.025)
Projection_results_n_85_h<-apply(Projection_results_n_85,2,quantile,0.975)
tt4<-cbind(c(2035:2100),Projection_results_n_85_m,Projection_results_n_85_l,Projection_results_n_85_h,rep("2Negative",66))

tt4<-data.frame(tt4)
colnames(tt4)<-c("year","median","lower","upper","type")
for(i in 1:4)
{
  tt4[,i]<-as.numeric(tt4[,i])
}

tt3<-rbind(tt3,tt4)

p2<-ggplot(data=tt3, aes(x=year, y=median, group=type,color=type))  
p2<-p2+geom_ribbon(aes(ymin=lower, ymax=upper,fill=type), colour=NA, alpha=0.6)+ geom_point() + geom_line()+scale_colour_manual(values = c("#44AA99","#DDCC77")) + scale_fill_manual(values = c("#44AA99","#DDCC77"))

grid.arrange (p,p2,nrow=2)

#concave vs convex
negative<-c()
for(i in 1:ncol(V))
{
  if(gam_majority2[i]==-2)
  {negative<-c(negative,i)}
}

V_future_s_neg<-V_future_s[,negative,]
V_pass_neg<-V[,negative]

V_future_s_neg_difference<-V_future_s_neg
for(k in 1:1848)
{
  V_future_s_neg_difference[k,,]<-V_future_s_neg[k,,]-t(V_pass_neg[,])
}
V_future_s_neg_difference_RCP45<-V_future_s_neg_difference[rcp=="RCP45",,]
V_future_s_neg_difference_RCP85<-V_future_s_neg_difference[rcp=="RCP85",,]

V_future_s_neg_value_RCP45<-V_future_s_neg[rcp=="RCP45",,]
V_future_s_neg_value_RCP85<-V_future_s_neg[rcp=="RCP85",,]

V_future_pass<-V_future_s_neg

Projection_results_diff<-matrix(data=NA,nrow=length(scenarios),ncol=66)
Projection_results_value<-matrix(data=NA,nrow=length(scenarios),ncol=66)
for(w in 1:length(scenarios))
{
  V_future_diff<-c()
  V_future_value<-c()
  for(j in ((w-1)*66+1):(w*66))
  {
     pass<-V_future_pass[j,,]
V_future_diff<-c(V_future_diff,mean((pass-t(V_pass_neg)),na.rm=TRUE))
V_future_value<-c(V_future_value,mean(pass,na.rm=TRUE))
}
Projection_results_diff[w,]<-V_future_diff
Projection_results_value[w,]<-V_future_value
}

Projection_results_diff_45<-Projection_results_diff[c(1,3:7,13,15,17,19:23),]
Projection_results_diff_85<-Projection_results_diff[c(2,8:12,14,16,18,24:28),]
Projection_results_value_45<-Projection_results_value[c(1,3:7,13,15,17,19:23),]
Projection_results_value_85<-Projection_results_value[c(2,8:12,14,16,18,24:28),]

Projection_results_diff_45_m<-apply(Projection_results_diff_45,2,mean,na.rm=TRUE)
Projection_results_diff_45_l<-apply(Projection_results_diff_45,2,quantile,0.025,na.rm=TRUE)
Projection_results_diff_45_h<-apply(Projection_results_diff_45,2,quantile,0.975,na.rm=TRUE)
tt5<-cbind(c(2035:2100),Projection_results_diff_45_m,Projection_results_diff_45_l,Projection_results_diff_45_h,rep("2Negative",66))

tt5<-data.frame(tt5)
colnames(tt5)<-c("year","median","lower","upper","type")
for(i in 1:4)
{
  tt5[,i]<-as.numeric(tt5[,i])
}

Projection_results_diff_85_m<-apply(Projection_results_diff_85,2,mean,na.rm=TRUE)
Projection_results_diff_85_l<-apply(Projection_results_diff_85,2,quantile,0.025,na.rm=TRUE)
Projection_results_diff_85_h<-apply(Projection_results_diff_85,2,quantile,0.975,na.rm=TRUE)
tt7<-cbind(c(2035:2100),Projection_results_diff_85_m,Projection_results_diff_85_l,Projection_results_diff_85_h,rep("2Negative",66))

tt7<-data.frame(tt7)
colnames(tt7)<-c("year","median","lower","upper","type")
for(i in 1:4)
{
  tt7[,i]<-as.numeric(tt7[,i])
}

positive<-c()
for(i in 1:ncol(V))
{
  if(gam_majority2[i]==2)
  {positive<-c(positive,i)}
}

V_future_s_pos<-V_future_s[,positive,]
V_pass_pos<-V[,positive]

V_future_s_pos_difference<-V_future_s_pos
for(k in 1:1848)
{
  V_future_s_pos_difference[k,,]<-V_future_s_pos[k,,]-t(V_pass_pos[,])
}
V_future_s_pos_difference_RCP45<-V_future_s_pos_difference[rcp=="RCP45",,]
V_future_s_pos_difference_RCP85<-V_future_s_pos_difference[rcp=="RCP85",,]

V_future_s_pos_value_RCP45<-V_future_s_pos[rcp=="RCP45",,]
V_future_s_pos_value_RCP85<-V_future_s_pos[rcp=="RCP85",,]

V_future_pass<-V_future_s_pos
Projection_results_diff<-matrix(data=NA,nrow=length(scenarios),ncol=66)
Projection_results_value<-matrix(data=NA,nrow=length(scenarios),ncol=66)
for(w in 1:length(scenarios))
{
  V_future_diff<-c()
  V_future_value<-c()
  for(j in ((w-1)*66+1):(w*66))
  {
     pass<-V_future_pass[j,,]
     V_future_diff<-c(V_future_diff,mean((pass-t(V_pass_pos)),na.rm=TRUE))
      V_future_value<-c(V_future_value,mean(pass,na.rm=TRUE))
}
Projection_results_diff[w,]<-V_future_diff
Projection_results_value[w,]<-V_future_value
}

Projection_results_diff_45<-Projection_results_diff[c(1,3:7,13,15,17,19:23),]
Projection_results_diff_85<-Projection_results_diff[c(2,8:12,14,16,18,24:28),]
Projection_results_value_45<-Projection_results_value[c(1,3:7,13,15,17,19:23),]
Projection_results_value_85<-Projection_results_value[c(2,8:12,14,16,18,24:28),]

Projection_results_diff_45_m<-apply(Projection_results_diff_45,2,mean,na.rm=TRUE)
Projection_results_diff_45_l<-apply(Projection_results_diff_45,2,quantile,0.025,na.rm=TRUE)
Projection_results_diff_45_h<-apply(Projection_results_diff_45,2,quantile,0.975,na.rm=TRUE)
tt6<-cbind(c(2035:2100),Projection_results_diff_45_m,Projection_results_diff_45_l,Projection_results_diff_45_h,rep("1Positive",66))

tt6<-data.frame(tt6)
colnames(tt6)<-c("year","median","lower","upper","type")
for(i in 1:4)
{
  tt6[,i]<-as.numeric(tt6[,i])
}

tt5<-rbind(tt5,tt6)

p3<-ggplot(data=tt5, aes(x=year, y=median, group=type,color=type))  
p3<-p3+geom_ribbon(aes(ymin=lower, ymax=upper,fill=type), colour=NA, alpha=0.6)+ geom_point() + geom_line()+scale_colour_manual(values = c("#44AA99","#DDCC77")) + scale_fill_manual(values = c("#44AA99","#DDCC77"))


Projection_results_diff_85_m<-apply(Projection_results_diff_85,2,mean,na.rm=TRUE)
Projection_results_diff_85_l<-apply(Projection_results_diff_85,2,quantile,0.025,na.rm=TRUE)
Projection_results_diff_85_h<-apply(Projection_results_diff_85,2,quantile,0.975,na.rm=TRUE)
tt8<-cbind(c(2035:2100),Projection_results_diff_85_m,Projection_results_diff_85_l,Projection_results_diff_85_h,rep("1Positive",66))

tt8<-data.frame(tt8)
colnames(tt8)<-c("year","median","lower","upper","type")
for(i in 1:4)
{
  tt8[,i]<-as.numeric(tt8[,i])
}

tt7<-rbind(tt7,tt8)

p4<-ggplot(data=tt7, aes(x=year, y=median, group=type,color=type))  
p4<-p4+geom_ribbon(aes(ymin=lower, ymax=upper,fill=type), colour=NA, alpha=0.6)+ geom_point() + geom_line()+scale_colour_manual(values = c("#44AA99","#DDCC77")) + scale_fill_manual(values = c("#44AA99","#DDCC77"))

grid.arrange (p3,p4,nrow=2)
```

## Summarizing the var x mean effects: past and future - Figure 3

```{r future , echo=FALSE}
# Mean, Var, Interaction
FutureMean45<-Mean_Var_future_s[c(1,3:7,13,15,17,19:23),,,1]
FutureMean85<-Mean_Var_future_s[c(2,8:12,14,16,18,24:28),,,1]
FutureVar45<-Mean_Var_future_s[c(1,3:7,13,15,17,19:23),,,2]
FutureVar85<-Mean_Var_future_s[c(2,8:12,14,16,18,24:28),,,2]
FutureInt45<-Mean_Var_future_s[c(1,3:7,13,15,17,19:23),,,3]
FutureInt85<-Mean_Var_future_s[c(2,8:12,14,16,18,24:28),,,3]

FutureMean45_m<-apply(FutureMean45,2,median)
FutureMean45_l<-apply(FutureMean45,2,quantile,0.025)
FutureMean45_h<-apply(FutureMean45,2,quantile,0.975)
tt_new<-cbind(FutureMean45_m,FutureMean45_l,FutureMean45_h,rep("1Mean",90))
FutureVar45_m<-apply(FutureVar45,2,median)
FutureVar45_l<-apply(FutureVar45,2,quantile,0.025)
FutureVar45_h<-apply(FutureVar45,2,quantile,0.975)
tt_new2<-cbind(FutureVar45_m,FutureVar45_l,FutureVar45_h,rep("2Var",90))
FutureInt45_m<-apply(FutureInt45,2,median)
FutureInt45_l<-apply(FutureInt45,2,quantile,0.025)
FutureInt45_h<-apply(FutureInt45,2,quantile,0.975)
tt_new3<-cbind(FutureInt45_m,FutureInt45_l,FutureInt45_h,rep("3Int",90))

tt_new<-rbind(tt_new,tt_new2,tt_new3)
tt_new<-data.frame(tt_new)
colnames(tt_new)<-c("median","lower","upper","type")
for(i in 1:3)
{
 tt_new[,i]<-as.numeric(as.character(tt_new[,i])) 
}

FutureMean85_m<-apply(FutureMean85,2,median)
FutureMean85_l<-apply(FutureMean85,2,quantile,0.025)
FutureMean85_h<-apply(FutureMean85,2,quantile,0.975)
tt_new4<-cbind(FutureMean85_m,FutureMean85_l,FutureMean85_h,rep("1Mean",90))
FutureVar85_m<-apply(FutureVar85,2,median)
FutureVar85_l<-apply(FutureVar85,2,quantile,0.025)
FutureVar85_h<-apply(FutureVar85,2,quantile,0.975)
tt_new5<-cbind(FutureVar85_m,FutureVar85_l,FutureVar85_h,rep("2Var",90))
FutureInt85_m<-apply(FutureInt85,2,median)
FutureInt85_l<-apply(FutureInt85,2,quantile,0.025)
FutureInt85_h<-apply(FutureInt85,2,quantile,0.975)
tt_new6<-cbind(FutureInt85_m,FutureInt85_l,FutureInt85_h,rep("3Int",90))

tt_new2<-rbind(tt_new4,tt_new5,tt_new6)
tt_new2<-data.frame(tt_new2)
colnames(tt_new2)<-c("median","lower","upper","type")
for(i in 1:3)
{
 tt_new2[,i]<-as.numeric(as.character(tt_new2[,i])) 
}


par(mfrow=c(1,1))

TEST<-c(apply(effects,2,median)[1],mean(tt_new[tt_new$type=="1Mean",1]),mean(tt_new2[tt_new2$type=="1Mean",1]),NA,apply(effects,2,median)[2],mean(tt_new[tt_new$type=="2Var",1]),mean(tt_new2[tt_new2$type=="2Var",1]),NA,apply(effects,2,median)[3],mean(tt_new[tt_new$type=="3Int",1]),mean(tt_new2[tt_new2$type=="3Int",1]))

TEST_reorder<-TEST[c(1,5,9,4,2,6,10,8,3,7,11)]

tt<-barplot(TEST_reorder,ylim=c(0,1),col=c("green","Blue","purple","white","green","Blue","purple","white","green","Blue","purple"))

TEST_L<-c(apply(effects,2,quantile,0.025)[1],mean(tt_new[tt_new$type=="1Mean",2]),mean(tt_new2[tt_new2$type=="1Mean",2]),NA,apply(effects,2,quantile,0.025)[2],mean(tt_new[tt_new$type=="2Var",2]),mean(tt_new2[tt_new2$type=="2Var",2]),NA,apply(effects,2,quantile,0.025)[3],mean(tt_new[tt_new$type=="3Int",2]),mean(tt_new2[tt_new2$type=="3Int",2]))

TEST_L_reorder<-TEST_L[c(1,5,9,4,2,6,10,8,3,7,11)]

TEST_H<-c(apply(effects,2,quantile,0.975)[1],mean(tt_new[tt_new$type=="1Mean",3]),mean(tt_new2[tt_new2$type=="1Mean",3]),NA,apply(effects,2,quantile,0.975)[2],mean(tt_new[tt_new$type=="2Var",3]),mean(tt_new2[tt_new2$type=="2Var",3]),NA,apply(effects,2,quantile,0.975)[3],mean(tt_new[tt_new$type=="3Int",3]),mean(tt_new2[tt_new2$type=="3Int",3]))

TEST_H_reorder<-TEST_H[c(1,5,9,4,2,6,10,8,3,7,11)]

segments(as.numeric(tt),TEST_L_reorder,as.numeric(tt),TEST_H_reorder)
```

